---
title: "Pushing the AR test beyond the limit"
subtitle: "Statlog heart disease example"
author: "Markku Kuismin"
institute: "University of Oulu, Northern Finland Birth Cohorts"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{css, echo = FALSE}
/* Define your custom class and font size here */
.small-font-slide {
  font-size: 90%;
}
```
---

class: center, middle

# A Statlog heart disease example

---
## A Statlog heart disease example

- The heart data exhibit features inconsistent with a smooth multivariate Gaussian model, for example:

  - Rounded blood-pressure measurements.
  - Cholesterol outliers.
  - Latent subgroup structure(?) 
  
- The AR gof test rejected multivariate normality, whereas the energy test doesn't

- The departure from normality is subtle and localized rather than a large-scale geometric discrepancy.

---
class: small-font-slide
## A Statlog heart disease example

```{r xaringan-fit-screen, echo=FALSE}
xaringanExtra::use_fit_screen()
```

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
library(tidyverse)

source("functions/read.libsvm.R")
source("functions/ar_dist.R")

set.seed(1)
```

```{r}
data_raw = read.libsvm("data/heart.txt",
                       dimensionality = 13)
```


1. Age: Patients Age in years (Numeric)
2. Sex: Gender (Male : 1; Female : 0) (Nominal)
3. cp: Type of chest pain experienced by patient. This term categorized into 4 category (0 typical angina, 1 atypical angina, 2 non- anginal pain, 3 asymptomatic (Nominal))
4. trestbps: patient's level of blood pressure at resting mode in mm/HG (Numerical)
5. chol: Serum cholesterol in mg/dl (Numeric)
6. fbs: Blood sugar levels on fasting > 120 mg/dl represents as 1 in case of true and 0 as false (Nominal)
7. restecg: Result of electrocardiogram while at rest are represented in 3 distinct values
(0 : Normal 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of >
0.05 mV) 2: showing probable or definite left ventricular hypertrophyby Estes' criteria (Nominal))
8. thalach: Maximum heart rate achieved (Numeric)
9. exang: Angina induced by exercise 0 depicting NO 1 depicting Yes (Nominal)
10. oldpeak: Exercise induced ST-depression in relative with the state of rest (Numeric)
11. slope: ST segment measured in terms of slope during peak exercise (0: up sloping; 1: flat; 2: down sloping(Nominal)
12. ca: The number of major vessels (0–3)(nominal))
13. thal: A blood disorder called thalassemia (0: NULL 1: normal blood flow 2: fixed defect (no blood flow in some part of the heart) 3: reversible defect (a blood flow is observed but it is not normal(nominal))
14. target: It is the target variable which we have to predict 1 means patient is suffering from heart disease and 0 means patient is normal.

```{r echo=FALSE}
y = data_raw[, 1]
X = data_raw[, -1]
colnames(X) <- c(
  "age",
  "sex",
  "cp",
  "trestbps",
  "chol",
  "fbs",
  "restecg",
  "thalach",
  "exang",
  "oldpeak",
  "slope",
  "ca",
  "thal"
)
Y = scale(X)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
d = ar_dist(x = Y, sim_dist = TRUE)

p = table(d$rho_sim)

dd = as.numeric(names(p))
dd = round(dd, 3)

df = data.frame(Prob = as.vector(p)/sum(as.vector(p)),
                rho = as.factor(dd))
```

---
## A Statlog heart disease example

- First, examine the whole data that includes several nominal features.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(df) + 
  geom_bar(aes(x = rho, y = Prob), stat="identity") + 
  geom_vline(
    xintercept = 1, 
    linewidth = 1,
    linetype = 2) +
  theme(
    axis.text.x = element_text(angle = 45)
  )
```

---
class: small-font-slide
## A Statlog heart disease example

- AR test statistic and interval estimates,

```{r}
d$rho
HDInterval::hdi(d$rho_sim)
```

- Energy test,

```{r}
energy::mvnorm.test(Y, R = 1000)
```

---
## A Statlog heart disease example

- Pick only continuous variables: age, resting blood pressure, cholesterol, maximum heart rate, oldpeak (ST depression).

```{r echo=FALSE, warning=FALSE, message=FALSE}
Y = X[, c(1, 4, 5, 8, 10)]
Y = scale(Y)

d = ar_dist(x = Y, sim_dist = TRUE)

p = table(d$rho_sim)

dd = as.numeric(names(p))
idx = which.min(abs(dd - d$rho)) + 1
dd = round(dd, 3)

df = data.frame(Prob = as.vector(p)/sum(as.vector(p)),
                rho = as.factor(dd))

gp = ggplot(df) + 
  geom_bar(aes(x = rho, y = Prob), stat="identity") + 
  geom_vline(
    xintercept = idx, 
    linewidth = 1,
    linetype = 2,
    color = "red") +
  theme(
    axis.text.x = element_text(angle = 45)
  )
#gp
```

---
class: small-font-slide
## A Statlog heart disease example

```{r eval=TRUE, message=FALSE, warning=FALSE}
d$rho
HDInterval::hdi(d$rho_sim)
```

```{r}
energy::mvnorm.test(Y, R = 1000)
```

---
## A Statlog heart disease example

- The variable 'oldpeak' is right skewed. First, remove it from the analysis,

```{r eval=TRUE, message=FALSE, warning=FALSE, fig.height=5.5}
plot(as.data.frame(Y))
```

---
## A Statlog heart disease example

```{r eval=TRUE, message=FALSE, warning=FALSE, echo =  FALSE}
Y = X[, c(1, 4, 5, 8)]
Y = scale(Y)

d = ar_dist(x = Y, sim_dist = TRUE)

p = table(d$rho_sim)

dd = as.numeric(names(p))
idx = which.min(abs(dd - d$rho)) + 1
dd = round(dd, 3)

df = data.frame(Prob = as.vector(p)/sum(as.vector(p)),
                rho = as.factor(dd))

gp = ggplot(df) + 
  geom_bar(aes(x = rho, y = Prob), stat="identity") + 
  geom_vline(
    xintercept = idx, 
    linewidth = 1,
    linetype = 2) +
  theme(
    axis.text.x = element_text(angle = 45)
  )
#gp
```

---
## A Statlog heart disease example

```{r eval=TRUE, message=FALSE, warning=FALSE}
d$rho
HDInterval::hdi(d$rho_sim)
```

```{r}
energy::mvnorm.test(Y, R = 1000)
```

---
## A Statlog heart disease example

- Use nonparanormal transformation to make the marginal distributions approximately standard normal,

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
Y = huge::huge.npn(Y, verbose = FALSE)

d = ar_dist(x = Y, sim_dist = TRUE)

p = table(d$rho_sim)

dd = as.numeric(names(p))
idx = which.min(abs(dd - d$rho)) + 1
dd = round(dd, 3)

df = data.frame(Prob = as.vector(p)/sum(as.vector(p)),
                rho = as.factor(dd))

gp = ggplot(df) + 
  geom_bar(aes(x = rho, y = Prob), stat="identity") + 
  geom_vline(
    xintercept = idx, 
    linewidth = 1,
    linetype = 2,
    color = "red") +
  theme(
    axis.text.x = element_text(angle = 45)
  )
#gp
```

---
## A Statlog heart disease example

```{r eval=TRUE, message=FALSE, warning=FALSE}
d$rho  
HDInterval::hdi(d$rho_sim)
```

```{r}
energy::mvnorm.test(Y, R = 1000)
```

---
## A Statlog heart disease example

- The resulting partial correlation matrix is

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
library(huge)

res = huge(x = Y, 
           nlambda = 100,
           method = "glasso",
           verbose = FALSE)

res_sel = huge.select(res, 
                      criterion = "ric",
                      verbose = FALSE)

R = -cov2cor(res_sel$opt.icov)
colnames(R) = rownames(R) = colnames(Y)

round(R, 3)
```

---
## A Statlog heart disease example

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
A = res_sel$opt.icov
A[A != 0] = 1
diag(A) = 0
colnames(A) = colnames(Y)
rownames(A) = colnames(Y)

G = igraph::graph_from_adjacency_matrix(A,
                                        mode = "undirected")
```

- The GGM estimated with Glasso is

```{r}
plot(G)
```

---
## A Statlog heart disease example

- The p-value of the AR test can be computed, e.g., using MC-testing,

```{r eval=TRUE, message=FALSE, warning=FALSE}
M = 10^3

rho_null = rep(0, M)

n = nrow(Y)

mu = rep(0, ncol(Y))

for(i in 1:M){
  
  XX = mvtnorm::rmvnorm(n = n, mean = mu)
  
  rho_null[i] = ar_dist(x = XX)$rho
  
}

(sum(rho_null <= d$rho) + 1)/(M + 1)
```

```{r eval=FALSE, echo=FALSE}
hist(rho_null, probability = TRUE)
abline(v = d$rho, lty = 2, lwd = 2)
```

---
## A Statlog heart disease example

```{r echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}
f = function(x) ar_dist(x)$rho

p = ncol(Y)

bootobj <- boot::boot(Y, 
                      statistic = f, 
                      R = M, 
                      sim = "parametric", 
                      ran.gen = function(x, y){
                        return(matrix(rnorm(n*p), 
                                      nrow = n, ncol = p))
                      })

mean(bootobj$t < bootobj$t0)
```

---

## A Statlog heart disease example

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.height=5.5}
data_long = as.data.frame(Y) %>% 
  pivot_longer(colnames(Y)) %>% 
  as.data.frame()

ggp <- ggplot(data_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density))) + 
  geom_density(col = "#1b98e0", linewidth = 2) + 
  facet_wrap(~ name, scales = "free")
ggp
```

- trestbps: Patient's level of blood pressure at resting mode in mm/HG

- Many repeated values -> Evidence against $H_0$.

---
## A Statlog heart disease example

- Use only features age, chol, and thalach

```{r eval=TRUE, message=FALSE, warning=FALSE}
Y = X[, c("age", "chol", "thalach")]
Y = huge::huge.npn(Y, verbose = FALSE)

d = ar_dist(Y, sim_dist = TRUE)

d$rho
HDInterval::hdi(d$rho_sim)
```

---
## A Statlog heart disease example

- After this, the data seems somewhat normal.

```{r eval=TRUE, message=FALSE, warning=FALSE}
(sum(rho_null < d$rho) + 1)/(M + 1)
energy::mvnorm.test(Y, R = M)
```
